---
title: "Assignment5A"
format: html
editor: visual
---

## Approach

For this assignment i am going to attempt to leverage mySQL and create a database with the data given. I think it will be good practice and good experience. Alot of this assignment will be learning the functions from tidyverse on how to clean up data. I will leverage LLM where needed to possibly explore options from tidyverse and see if there is a list of functions. I find using LLM to describe best practices to problems is also a good way to use the LLM.

```{r}
library(tidyverse)
```

```{r}
airlines_data <- "airline,status,Los_Angeles,Phoenix,San_Diego,San_Francisco,Seattle
ALASKA,on time,497,221,212,503,1841
ALASKA,delayed,62,12,20,102,305
AM WEST,on time,694,4840,383,320,201
AM WEST,delayed,117,415,65,129,61"

write.csv(read.csv(text = airlines_data), "airlines.csv", row.names = FALSE)

airlines <- read.csv("airlines.csv")
airlines

```

```{r}
airlines_tidy <- airlines %>%
  pivot_longer(
    cols = c(Los_Angeles, Phoenix, San_Diego, San_Francisco, Seattle),
    names_to = "destination",
    values_to = "flights"
  )
airlines_tidy
```

```{r}
airlines_wide <- airlines_tidy %>%
  pivot_wider(
    names_from = status,
    values_from = flights
  ) %>%
  rename(on_time = `on time`) %>%
  mutate(
    total = on_time + delayed,
    delay_rate = round(delayed / total * 100, 1)
  )
airlines_wide
```

```{r}
overall <- airlines_wide %>%
  group_by(airline) %>%
  summarise(
    total_flights = sum(total),
    total_delayed = sum(delayed),
    overall_delay_rate = round(total_delayed / total_flights * 100, 1)
  )
overall
```

```{r}
ggplot(airlines_wide, aes(x = destination, y = delay_rate, fill = airline)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Delay Rate by Destination and Airline",
    x = "Destination",
    y = "Delay Rate (%)",
    fill = "Airline"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "tomato"))
```

## Conclusions

After tidying and transforming the data, the analysis reveals an interesting paradox. On the surface, AM WEST appears to have a lower overall delay rate than ALASKA. However, when broken down by destination, ALASKA actually outperforms AM WEST at every single city.
